Dissecting the causal question underlying the association between cancer and ADRD: an opportunity to understand sources of bias

### 2021-05-31 04_reports/manuscript.Rmd
updates and more updates
deleted old stuff, added refs, add bibtex

---
title: "Dissecting the causal question underlying the association between cancer and ADRD "
author: "L. Paloma Rojas-Saunero"
date: ""
output: [html_document]
editor_options: 
  chunk_output_type: console
---
```{r, echo =  FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(survival)
library(tidyverse)
library(tableone)
library(magrittr)
library(kableExtra)
# source(here::here("02_R", "03_descriptives.R"))
# source(here::here("02_R", "05_plots.R"))
# source(here::here("02_R", "05b_plot_causes_death.R"))
# table_results <- rio::import(here::here("02_R", "table_results.csv"))
# table_results_bounds <- rio::import(here::here("02_R", "table_results_bounds.csv"))
# models <- read_csv(here::here("04_reports", "models.csv")) %>% select(-6)
```

@@

Many observational studies have consistently found that individuals with cancer have a lower risk of developing of Alzheimer´s disease or related dementias (ADRD) when compared to individuals with no history of cancer1–4.  These findings have motivated substantial research toward mechanistic explanations, including searching for and hypothesizing that molecular and genetic mechanisms may explain this association5–12. These research inquiries inevitably lead to discussions of repurposing or augmenting current cancer chemotherapeutics for ADRD13.

Nevertheless, inferring any treatment or mechanistic effects from the observed cancer-ADRD inverse association is not straightforward. Researchers have raised concerns related to the competing event of death, unmeasured confounding, and ascertainment error that could explain these results9,14. However, understanding these or other sources of bias first requires that we make explicit the causal question. Moreover, making explicit the causal question is one step toward tying a research study to a question that is relevant to decision-making.

To illustrate the complexities of inferring hypothetical or available treatments’ effects on ADRD from the observed cancer-ADRD association, we focus on the   Pin1 enzyme. Previous animal studies have shown that the Pin1 enzyme over-expression promotes tumorigenesis, while its down-regulation is attributed to mechanisms that contribute to neurodegeneration and amyloid deposition11,12,15. If we one day could develop a drug that increases Pin1 expression specifically in brain tissue in hopes of preventing dementia, we could pose the question as: _What is the direct effect of this Pin1-targeting drug on the risk of ADRD over time compared to standard treatments?_

To explore how we might learn about this effect using real-world data on cancer and ADRD, we progressively build a causal directed acyclic graph to connect this particular causal question to the observable data and the assumptions we rely on to identify the effect.  We exemplify different scenarios with data collected from the Rotterdam Study, a population-based cohort study. We describe the challenges and how they translate into the analytic decisions. Last, we discuss how information on mortality and cause of death can provides insight about the direction of some sources of bias.

@@

If this hypothetical Pin1-targeting drug was developed, the best way to understand its effect on dementia risk would be to do a well-conducted randomized trial in which we randomize eligible participants in late midlife (e.g., ages 50-60 years) to receive this drug or not, and closely monitor ADRD over a lengthy follow-up.

Since there is no drug currently available that targets Pin1, at best we can use observational data on Pin 1 expression measurements. For example, suppose that a biomarker test was available to measure Pin1 and we measured this biomarker from (stored) baseline blood samples in a population based-cohort that recruited participants in late midlife. Since the biomarker Pin1 is measured within an observational study, confounding can explain an observed association between it and ADRD. In Figure 1, we show that, Pin1 expression ($P_0$) and ADRD at time $t+1$ ($Y_{t+1}$) may share causes $L$, and that assessing the causal relationship requires adjusting for these confounders $L$. Previous studies have described age, sex, educational level and race/ethnicity as the minimal adjusting set of covariates (Ospina). However, environmental and behavioral factors such as smoking, which are known to cause microenvironmental changes such as inflammation and changes in tissue remodeling, may translate into Pin1 over-expression and are also related to the development of ADRD.

For simplicity, we treat Pin1 expression as a point intervention and we fix the time-ordering of covariates (that is, we assume $L$ happens prior to $P_0$). In reality, it is possible that Pin1 expression changes over time and is affected by time-varying confounders (like smoking), which produces treatment-confounder feedback loops. Addressing such time-varying confounding would require repeated measurements of  $L$ and $P_0$16.

```{tikz, echo = FALSE, fig.cap = "Pin1 as biomarker and risk of dementia"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [right of = p, gray!30] (dp) {$D_t$};
\node [right of = dp, gray!30] (pstar) {$P^*_t$};
\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C$};
\draw[->] (l) -- (p);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->, gray!30] (d) -- (y);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\draw[->, semithick] (p) to [out=45,in=135, looseness=0.2] (y);
\end{tikzpicture}
```

However, unfortunately, in current studies we do not even have a single measurement of Pin1 expression, let alone repeated measurements. Thus, we can only rely on a proxy of this exposure. Since Pin1 over-expression is present in tumors, and tumors are only measured through screening and diagnosis, we considered cancer diagnosis as the proxy for Pin1 over-expression4,17–24. We depict this feature in Figure 2, where $P^*$ represents _incident cancer diagnosis_, the measured proxy of $P_0$. In this causal graph  we colored the path $P^* \leftarrow P \leftarrow Y_{t+1}$ because although we are measuring the association between $P^*$ and $Y_{t+1}$ in the observed data, we are assuming that the captured effect is only through $P_0$.

```{tikz, echo = FALSE, fig.cap = "Cancer diagnosis as proxy for Pin1 expression"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm, gray!30](l) {$L$};
\node [right of = p, gray!30] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P^*_t$};
\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C$};
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->, gray!30] (d) -- (y);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

A major challenge related to assigning cancer diagnosis as the proxy of Pin1 is defining time zero. Not everyone who had Pin1 over-expression will be diagnosed with cancer by late-midlife, but in late-life. For this reason prospective cohort studies have considered cancer diagnosis as a time-varying or time-dependent exposure. This means that a participant within a prospective cohort study contributes to the "regulated Pin1" arm since study entry up to the time of cancer diagnosis and later on to the "Over-expressed Pin1" arm. Other studies have included participants with cancer diagnosis at the time of the diagnosis (for example from cancer registries19,23,25,26) and matched participants by age. In both cases we must remember that the main interest is on the unmeasured $P_0$ thus we should only adjust for covariates prior to $P_0$ and be careful to adjust for post-baseline covariates of $P_0$ or mediators between $P_0$ and $P^*$.

One of the mediators between $P_0$ and $P^*$ is death prior to cancer diagnosis, represented as $D_t$ in Figure 3. Since individuals are at risk of dying from other causes (such as cardiovascular death), we can only measure $P^*$ in the subset of individuals who have survived long enough to have a cancer diagnosis. Several risk factors that increase the risk of cancer might also cause death prior to cancer diagnosis, for example smoking may cause lung cancer and chronic obstructive pulmonary disease (a leading cause of death in this age group27). Therefore, we can only isolate the effect of $P_0$ in $P^*$ if we condition on $C_1$ and block the backdoor pathway that is open by conditioning on $D_t$. In this way we assume a hypothetical scenario in which we could prevent death prior to cancer diagnosis, by conditioning on a rich set of covariates28. This assumption must hold regardless of whether we use incident cancer as a time-varying exposure, or match cancer patients to participants free of cancer by age.

Furthermore, we note that this is only one of the issues with considering cancer diagnosis as the proxy for Pin1 expression in terms of information bias(ref). We could add more complexity by considering further shared causes of cancer and ADRD diagnosis, or further mediators between Pin1 expression and cancer diagnosis, including but not limited to: screening guidelines, type of healthcare coverage, health-seeking behaviors, and healthcare availability.

```{tikz, echo = FALSE, fig.cap = "Survival bias when treating cancer diagnosis as the proxy for Pin1 expression"}

\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm, gray!30](l) {$L$};
\node [right of = p] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P^*_t$};
\node [right of = pstar, gray!30] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = d, yshift = 1cm, xshift = 1cm, gray!30](cd) {$C$};
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->] (dp) -- (pstar);
\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->] (d) -- (y);
\draw[->, gray!30] (cd) -- (d);
\draw[->, gray!30] (cd) -- (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

Up to this point we have outlined certain underlying assumptions of using incident cancer diagnosis as a proxy for Pin1 expression. However, as mortality increases steeply in late life, even in the setting of the ideal randomized trial of a hypothetical Pin1-targeting drug, we can only measure ADRD over follow-up in the individuals who survive long enough to have a diagnosis. For this reason, death is a competing event of ADRD because if a participant dies prior to ADRD diagnosis, death prevents observing ADRD at future time-points. As such, to identify the direct effect of $P_0$ in $Y_{t+1}$ we have to first assess if the drug has an any side effects that could increase the risk of death in any way. If the drug has no systemic side-effects, such as that there is no arrow between $P_0$ and $D_{t+1}$, such as observed in Figure 4.

```{tikz, echo = FALSE, fig.cap = "Death as a competing event for dementia over follow-up"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm, gray!30](l) {$L$};
\node [right of = p, gray!30, gray!30] (dp) {$D_t$};
\node [right of = dp, gray!30] (pstar) {$P^*_t$};
\node [right of = pstar] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = d, yshift = 1cm, xshift = 1cm,](cd) {$C$};
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->, gray!30] (pstar) -- (d);
\draw[->] (d) -- (y);
\draw[->] (cd) -- (d);
\draw[->] (cd) -- (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

However, if the drug has any side-effects we have to decide if either we want to estimate a total effect (that is, the effect on ADRD even if partly mediated by the effect on death) or a direct effect (that is, the effect directly on ADRD not mediated by death) (ref). We face this challenge in our observational data setting, since we are measuring cancer diagnosis as the proxy for Pin1 over-expression, and cancer is a leading cause of death29. We can visualize this data feature in the causal diagram in Figure 5. In this causal diagram we include $D_{t+1}$ as a representation of death over follow-up, an arrow between $P_0$ and $D_{t+1}$ since $P_0$ may increase the mortality risk. We also include an arrow between $P^*_t$ and $D_{t+1}$ since cancer diagnosis, and subsequent treatment (or lack of treatment) may have an effect on death. Last, the arrow from $D_{t+1}$ and $Y_{t+1}$ represents the key feature of a competing events data structure: an individual who dies over follow-up (prior to dementia diagnosis) cannot subsequently develop ADRD, and since $D_{t+1}$ and $Y_{t+1}$ are events related to aging, $C$ represent the shared causes of both events.

In this causal diagram we observe that, had we measured and adjusted for $L$, we could estimate the total effect of $P_0$ in $Y_{t+1}$ without further assumptions. However the total effect includes all pathways between $P_0$ and $Y_{20}$28. This means that if Pin1 has an effect on mortality through cancer or other mechanisms, we will observe a protective effect of Pin1 over-expression in ADRD, partially or fully mediated through death. Since the question of interest is focused on the direct effect of $P_0$ in $Y_{t+1}$ as in Figure 4 (violet arrows) we need to conceptualize the different mechanisms through which $P_0$ could affect $D_{t+1}$ and $Y_{20}$. With this in mind we can conceive different causal questions (estimands) to represent this direct effect , such as the controlled direct effect (CDE) and the natural separable direct effect. In this section we discuss the controlled direct effect as the causal question of interest since it translates to frequently used methods in this literature (such as Kaplan-Meier estimator and Cox-proportional hazard model) and leave the separable direct effects question for discussion.

```{tikz, echo = FALSE, fig.cap = "Direct effect"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm, gray!30](l) {$L$};
\node [right of = p, gray!30, gray!30] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P^*_t$};
\node [right of = pstar] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = d, yshift = 1cm, xshift = 1cm](cd) {$C$};
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, gray!30] (dp) -- (pstar);
\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->] (pstar) -- (d);
\draw[->] (d) -- (y);
\draw[->] (cd) -- (d);
\draw[->] (cd) -- (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

The CDE represents the effect of Pin1 over-expression in ADRD in a setting where we could have prevented death over the study period. It relies on the assumption that we have measured all $C$ to block the pathway $Y20 \leftarrow C  \rightarrow D_{15} \rightarrow P \rightarrow P*$. This assumption is defined as the independent censoring assumption conditional on covariates. In this setting death is treated as a censoring event and it can be interpreted as those who died would have the same risk of developing dementia if prevented from dying than those who remained alive and free of dementia at a given time point, conditional on shared causes of dementia and death.

Therefore, if we combine the challenges in section two related to cancer diagnosis as a proxy for Pin1 over-expression, and having death as a competing event of ADRD we observe the complexity of the DAG in Figure 5. Though this is yet a simplified version since we omitted time varying $P*$, $L$ and $C's$.

```{tikz, echo = FALSE, fig.cap = "Direct effect"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P_0$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [right of = p] (dp) {$D_t$};
\node [right of = dp] (pstar) {$P^*_t$};
\node [right of = pstar] (d) {$D_{t+1}$};
\node [right of = d] (y) {$Y_{t+1}$};
\node [below of = d, yshift = 1cm, xshift = 1cm](cd) {$C$};
\draw[->] (l) -- (p);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->] (dp) -- (pstar);
\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar);
\draw[->] (pstar) -- (d);
\draw[->] (d) -- (y);
\draw[->] (cd) -- (d);
\draw[->] (cd) -- (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

>>>>

---
title: "Dissecting the causal question underlying the association between cancer and ADRD: an opportunity to understand sources of bias"
author: "L. Paloma Rojas-Saunero"
date: ""
output: [html_document]
editor_options: 
  chunk_output_type: console
bibliography: cancer_dementia.bib
---
```{r, echo =  FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(survival)
library(tidyverse)
library(tableone)
library(magrittr)
library(kableExtra)
# source(here::here("02_R", "03_descriptives.R"))
# source(here::here("02_R", "05_plots.R"))
# source(here::here("02_R", "05b_plot_causes_death.R"))
# table_results <- rio::import(here::here("02_R", "table_results.csv"))
# table_results_bounds <- rio::import(here::here("02_R", "table_results_bounds.csv"))
# models <- read_csv(here::here("04_reports", "models.csv")) %>% select(-6)
```

@@

Many observational studies have consistently found that individuals with cancer have a lower risk of developing of Alzheimer´s disease or related dementias (ADRD) when compared to individuals with no history of cancer [@ma2014;@hanson2016;@vanderwillik2018;@ospina2020] .  These findings have motivated substantial research toward mechanistic explanations, including searching for and hypothesizing that molecular and genetic mechanisms may explain this association[@behrens2009;@harris2014;@nudelman2019;@Papin2020;@driverbiogeront2014;@olson2019;@li2021;@driverpin12015]. These research inquiries inevitably lead to discussions of repurposing or augmenting current cancer chemotherapeutics for ADRD [@Snyder2017].

Nevertheless, inferring any treatment or mechanistic effects from the observed cancer-ADRD inverse association is not straightforward. Researchers have raised concerns related to the competing event of death, unmeasured confounding, and ascertainment error that could explain these results[@driverbiogeront2014;@ganguli2015]. However, understanding these or other sources of bias first requires making explicit the causal question. Moreover, making explicit the causal question is one step toward tying a research study to a question that is relevant to decision-making[@didelez2016;@labrecque2017].

To illustrate the complexities of inferring hypothetical or available treatments’ effects on ADRD from the observed cancer-ADRD association, we focus on a specific question conceptualizing the  Pin1 enzyme as the target of intervention. Previous animal studies have shown that Pin1 enzyme over-expression promotes tumorigenesis, while its down-regulation is attributed to mechanisms that contribute to neurodegeneration and amyloid deposition[@driverpin12015;@angelucci2017;@li2021]. If we one day could develop a drug that increases Pin1 expression specifically in brain tissue in hopes of preventing dementia, we could pose the question as: _What is the effect of this Pin1-targeting drug on the risk of ADRD over time compared to standard treatments?_

To explore how we might learn about this effect using real-world data on cancer and ADRD, we progressively build a causal directed acyclic graph to connect this particular causal question to the observable data and the assumptions we rely on to study the effect. We exemplify different scenarios with data collected from the Rotterdam Study, a population-based cohort study. We describe the challenges and how they translate into the analytic decisions. Last, we discuss how information on mortality and cause of death can provides insight about the direction of some sources of bias.

@@

If this hypothetical Pin1-targeting drug was developed, the best way to understand its effect on dementia risk would be to have a well-conducted randomized trial in which we randomize eligible participants in late midlife (e.g., ages 50-60 years) to receive this drug or not, and closely monitor ADRD over a lengthy follow-up. Since this drug is not currently available, at best we can use observational data on Pin1 expression measurements. For example, suppose that a biomarker test was available to measure Pin1 and we measured this biomarker from (stored) baseline blood samples in a population based-cohort that recruited participants in late midlife. Although this means we could measure the biomarker at a particular time in everyone, defined by the cohort study design, an unanswered question that would remain is related to the best time to measure the biomarker in an applied setting. This question is beyond the scope of this exercise and for didactic purpose we may consider that the study entrance (and subsequent data collection) date was part of a screening process.

In this observational study, confounding could explain the observed association between Pin1 and ADRD. In the directed acyclic graph [@whatif2020] of Figure 1, we present Pin1 expression as $P$ and ADRD diagnosis over time as $Y_{t+1}$. Both nodes may share causes $L$, and to assess the causal relationship between $P$ and $Y_{t+1}$ we would require adjusting for $L$. Previous studies have described age, sex, educational level and race/ethnicity as the minimal adjusting set of covariates[@ospina2020]. However, environmental and behavioral factors such as smoking, which are known to cause microenvironmental changes such as inflammation and changes in tissue remodeling, may translate into Pin1 over-expression and are also related to the development of ADRD[ref].

```{tikz, echo = FALSE, fig.cap = "Pin1 as biomarker and risk of dementia"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [gray!30, right of = p] (pstar) {$P^*_t$};
\node [gray!30, right of = pstar] (dp) {$D_t$};
\node [gray!30, right of = dp] (pstar2) {$P^*_{t+1}$};
\node [gray!30, right of = pstar2] (dp2) {$D_{t+1}$};
\node [right of = dp2] (y) {$Y_{t+1}$};
\node [gray!30, below of = dp2, yshift = 1cm](cd) {$C$};
\draw[->] (l) -- (p);
\draw[->, gray!30] (l) to (dp);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, gray!30] (p) to (pstar);
\draw[->, gray!30] (pstar) -- (dp);
\draw[->, gray!30] (dp) -- (pstar2);
\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar2);
\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[->, gray!30] (dp2) -- (y);
\draw[->, gray!30] (pstar2) -- (dp2);
\draw[->, gray!30] (cd) -- (dp);
\draw[->, gray!30] (cd) -- (dp2);
\draw[->, gray!30] (cd) -- (y);
\draw[->, gray!30, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

Currently, Pin1 expression is not an available biomarker for population-based research, so instead we may have to measure a proxy to approximate this exposure. Because Pin1 over-expression is present in tumors, and tumors are only measured through screening and diagnosis, previous studies have considered Pin1 over-expression as the explanation behind the inverse association between cancer and ADRD, though Pin1 was not explicitly part of the research question[@ospina2020;@driver2012;@musicco2013;@freedman2016;@bowles2017;@frain2017;@schmidt2017;@sun2020;@ording2020].

In a cohort setting, participants may have a cancer diagnosis at any point over follow-up. As such, we may consider cancer diagnosis as a time-varying process and to define cancer diagnosis as the proxy for Pin1 we may assume that Pin1 is over-expressed only by the time a person has the diagnosis, though this assumption may not represent the true biological process. We depict this feature in Figure 2, where $P^*_t$ and $P^*_{t+1}$ represent _cancer diagnosis_ over time, the measured proxy of $P$. Although we are measuring the association between cancer diagnosis over time and ADRD in the observed data, we are assuming that the captured effect is only through the pathway that involves $P$.

```{tikz, echo = FALSE, fig.cap = "Cancer diagnosis as proxy for Pin1 expression"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P$};
\node [gray!30, above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [right of = p] (pstar) {$P^*_t$};
\node [gray!30, right of = pstar] (dp) {$D_t$};
\node [right of = dp] (pstar2) {$P^*_{t+1}$};
\node [gray!30, right of = pstar2] (dp2) {$D_{t+1}$};
\node [right of = dp2] (y) {$Y_{t+1}$};
\node [gray!30, below of = dp2, yshift = 1cm](cd) {$C$};
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to (dp);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->] (p) to (pstar);
\draw[->, gray!30] (pstar) -- (dp);
\draw[->, gray!30] (dp) -- (pstar2);
\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar2);
\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[->, gray!30] (dp2) -- (y);
\draw[->, gray!30] (pstar2) -- (dp2);
\draw[->, gray!30] (cd) -- (dp);
\draw[->, gray!30] (cd) -- (dp2);
\draw[->, gray!30] (cd) -- (y);
\draw[->, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

Given the nature of an elderly cohort, cancer diagnosis will be measured in the subset of the population who survives other causes of death. In Figure 3 we represent death prior to cancer diagnosis as $D_t$, and the arrow between $D_t$ and $P^*_{t+1}$ represents a deterministic association such as that $P^*$ is only observed if $D_t$ is zero. The shared causes between $P$ and $D_t$ are also represented by $L$,  for simplicity we illustrate them with the same node, however, although some shared causes might affect $Y_{t+1}$ this representation would not restrict to those covariates only.

We assume that there is no effect of Pin1 over-expression in all-cause mortality prior to cancer diagnosis (no arrow between $P$ and $D_t$), however dying prior to cancer diagnosis may represent a selection mechanism that prevents generalizability or transportability of the effect measure to the entire target population [@hernan2017].

Furthermore, we could consider unmeasured and differential measurement error due to other shared causes of cancer and ADRD including but not limited to: screening guidelines, type of health care coverage, health-seeking behaviors, and health care availability.

```{tikz, echo = FALSE, fig.cap = "Death prior to cancer diagnosis"}

\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [right of = p] (pstar) {$P^*_t$};
\node [right of = pstar] (dp) {$D_t$};
\node [right of = dp] (pstar2) {$P^*_{t+1}$};
\node [gray!30, right of = pstar2] (dp2) {$D_{t+1}$};
\node [right of = dp2] (y) {$Y_{t+1}$};
\node [gray!30, below of = dp2, yshift = 1cm](cd) {$C$};
\draw[->] (l) -- (p);
\draw[->] (l) to (dp);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->] (p) to (pstar);
\draw[->, gray!30] (pstar) -- (dp);
\draw[->] (dp) -- (pstar2);
\draw[->] (p) to [out=45,in=135, looseness=0.3] (pstar2);
\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[->, gray!30] (dp2) -- (y);
\draw[->, gray!30] (pstar2) -- (dp2);
\draw[->, gray!30] (cd) -- (dp);
\draw[->, gray!30] (cd) -- (dp2);
\draw[->, gray!30] (cd) -- (y);
\draw[->, gray!30, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

Even if survival prior to cancer diagnosis does not introduce bias over time, we can only measure ADRD over follow-up in the individuals who survive long enough to have an ADRD diagnosis, as represented in Figure 4. In this causal diagram we include one more node of death over time that comes after the exposure, represented as $D_{t+1}$, this acts as a competing event of $Y_{t+1}$ because if a participant dies prior to ADRD diagnosis, death prevents observing ADRD at future time-points. Furthermore, since $D_{t+1}$ and $Y_{t+1}$ are events related to aging and its consequences, $C$ represents the shared causes of both events such as cardiovascular conditions.

```{tikz, echo = FALSE, fig.cap = "Death as a competing event for dementia over follow-up"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P$};
\node [above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [right of = p, gray!30] (pstar) {$P^*_t$};
\node [right of = pstar] (dp) {$D_t$};
\node [right of = dp, gray!30] (pstar2) {$P^*_{t+1}$};
\node [right of = pstar2] (dp2) {$D_{t+1}$};
\node [right of = dp2] (y) {$Y_{t+1}$};
\node [below of = dp2, yshift = 1cm](cd) {$C$};
\draw[->, gray!30] (pstar2) -- (dp2);
\draw[->, gray!30] (p) to (pstar);
\draw[->, gray!30] (pstar) -- (dp);
\draw[->, gray!30] (dp) -- (pstar2);
\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar2);
\draw[->, gray!30, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\draw[->] (l) -- (p);
\draw[->] (l) to (dp);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[->] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[->] (dp2) -- (y);
\draw[->] (cd) -- (dp);
\draw[->] (cd) -- (dp2);
\draw[->] (cd) -- (y);
\end{tikzpicture}
```

In the scenario where $P$ represented the targeted-drug for Pin1, and if this drug has no systemic beneficial or harmful side-effects such as that there is no arrow between $P$ and $D_{t+1}$, a total effect would quantify the effect of $P_0$ on $Y_{t+1}$ that does not include any pathway mediated through $D_{t+1}$[@young2020].

However, in the context of cancer diagnosis we cannot rule out it's effect on death, as observed in Figure 5. In this case a total effect of $P$ in $Y_{t+1}$ would include the causal pathway mediated by the effect of cancer diagnosis in mortality, which may translate into an inverse association[29]. To isolate the direct effect of $P$ in $Y_{t+1}$ through measurement of $P^*_t$ we have at least two alternatives. We can either imagine a causal question where we decompose the effect of cancer by the different mechanisms that affect dementia and death separately[@stensrud2020]. Alternatively, we can define an scenario where death could have been prevented. The latter is defined as the controlled direct effect, where death is considered as a censoring event and it relies on the assumption that we have measured all $C$ to block the pathway $Y_{t+1} \leftarrow C  \rightarrow D_{t+1} \rightarrow P^*_t \rightarrow P_0$. Previous studies have defined death as a censoring event [ref], although failed to explicitly define how to relax the independent censoring assumption and while adjusting for confounders for dementia may represent partially the adjustment for the competing event of death [@young2020], adjusting for baseline covariates may not be sufficient. Furthermore interpretation regarding the prevention of death was not explicitly portrait.

```{tikz, echo = FALSE, fig.cap = "Direct effect"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2cm]
\node (p) {$P$};
\node [gray!30, above of = p, yshift = -0.8cm, xshift = -1cm](l) {$L$};
\node [gray!30, right of = p] (pstar) {$P^*_t$};
\node [gray!30, right of = pstar] (dp) {$D_t$};
\node [gray!30, right of = dp] (pstar2) {$P^*_{t+1}$};
\node [gray!30, right of = pstar2] (dp2) {$D_{t+1}$};
\node [gray!30, right of = dp2] (y) {$Y_{t+1}$};
\node [gray!30, below of = dp2, yshift = 1cm](cd) {$C$};
\draw[->, gray!30] (l) -- (p);
\draw[->, gray!30] (l) to (dp);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (dp2);
\draw[->, gray!30] (l) to [out=0,in=90, looseness=0.4] (y);
\draw[->, gray!30] (p) to (pstar);
\draw[->, gray!30] (pstar) -- (dp);
\draw[->, gray!30] (dp) -- (pstar2);
\draw[->, gray!30] (p) to [out=45,in=135, looseness=0.3] (pstar2);
\draw[->, gray!30] (dp) to [out=-45,in=-135, looseness=0.3] (dp2);
\draw[->, gray!30] (dp2) -- (y);
\draw[->, gray!30] (pstar2) -- (dp2);
\draw[->, gray!30] (cd) -- (dp);
\draw[->, gray!30] (cd) -- (dp2);
\draw[->, gray!30] (cd) -- (y);
\draw[->, gray!30, semithick] (p) to [out=+45,in=+135, looseness=0.3] (y);
\end{tikzpicture}
```

Finally, if we combine the challenges related to cancer diagnosis as a proxy for Pin1 over-expression, and having death as a competing event of ADRD we observe the complexity of the causal diagram in Figure 5. This As mentioned above, this is nevertheless a simplified version since we omitted the time varying nature of all nodes in the graph, and thus are ignoring how feedback loops between them further complicate interpretability and identificability. We also ignore further measurement error on $P^*_t$ and $Y_{t+1}$.

We now turn to an application where we show how these challenges, tied to different assumptions lead to different analytic decisions, and we will show ways to mitigate or better understand them given current available data.  Nevertheless, we acknowledge that given the large gap between $P_0$ and $P^*_t$, not only in terms of time between measurements and potential selection bias, but since cancer diagnosis represents a complex and heterogenous health condition that exceeds the representation of Pin1 expression, at best we may test the sharp null hypothesis rather than estimating the causal effect of Pin1.
